---
title: "Example 4: Nest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example4_nest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
  This vignette displays how to use nesting in `tidyfst`. It has referred to `tidyr`s vignette in <https://tidyr.tidyverse.org/articles/nest.html>. Now fist, we nest the "mtcars" data.frame by "cyl" column.
```{r setup}
library(tidyfst)

# nest by "cyl" column
mtcars_nested <- mtcars %>% 
  nest_by("cyl")

# inspect the output data.table
mtcars_nested

```

  Now, we want to do a regression within the nested group "cyl". We'll use the famous `lapply` to complete this:

```{r}
mtcars_nested2 <- mtcars_nested %>% 
  mutate_dt(model = lapply(ndt,function(df) lm(mpg ~ wt, data = df)))

mtcars_nested2
```
  We could see that the model is stored in the column "model".
  Now, we try to get the fitted value in the model.
```{r}
mtcars_nested3 <- mtcars_nested2 %>% 
  mutate_dt(model_predict = lapply(model, predict))
mtcars_nested3$model_predict
```
  We could find that the "model_predict" is a list of numeric vectors. Let's try to unnest the target column "model_predict". 
```{r}
mtcars_nested3 %>% unnest_col(model_predict)
```
  This process would remove all the other list column automatically. For instance, in our case, the column "ndt" is removed.
