---
title: "Example 3: Reshape"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 3: Reshape}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
  When I design `longer_dt` and `wider_dt`, I could find the `pivot_longer` and `pivot_wider` in `tidyr` and `melt` and `dcast` in `data.table`. Still, designing this API is not easy, as my goal is to let users use it with least pain. Here we would try to reproduce the results in the vignette of `tidyr`(<https://tidyr.tidyverse.org/articles/pivot.html>). First load the packages:
```{r setup}
library(tidyfst)
library(tidyr)
```

## Longer
  First inspect the data:
```{r}
relig_income
```
  In `tidyr`, to get the longer format you need:
```{r,eval=FALSE}
relig_income %>% 
  pivot_longer(-religion, names_to = "income", values_to = "count")
```
  In `tidyfst`, we have:
```{r,warning=FALSE}
relig_income %>% 
  longer_dt("religion",name = "income",value = "count")
```
  Another example from `tidyr`:
```{r,eval=FALSE}
billboard

# tidyr way:
 billboard %>%
   pivot_longer(
     cols = starts_with("wk"),
     names_to = "week",
     values_to = "rank",
     values_drop_na = TRUE
   )

# tidyfst way:
billboard %>% 
  longer_dt(-"wk",
            name = "week",
            value = "rank",
            na.rm = TRUE
            )
# regex could select the groups to keep, and minus could select the reverse
```
A warning would could come out because the merging column has different data types and do the coercion automatically.
    
## Wider

```{r}
## data
fish_encounters

## tidyr way:
fish_encounters %>% 
  pivot_wider(names_from = station, values_from = seen)

## tidyfst way:
fish_encounters %>% 
  wider_dt(name = "station",value = "seen")

# if no keeped groups are selected, use all except for name and value columns
```
    
  If you want to fill with 0s, use:
```{r}
fish_encounters %>% 
  wider_dt(name = "station",value = "seen",fill = 0)
```
  Note that the parameter of `name` and `value` should always be provided and should be explicit called (with the parameter names attached).

## More complicated example
This example comes from data.table (<https://rdatatable.gitlab.io/data.table/articles/datatable-reshape.html>), and has been used in tidyr too. We'll try to do it in tidyfst in this example.
If we have a data.frame as below:
```{r}
family <- fread("family_id age_mother dob_child1 dob_child2 dob_child3 gender_child1 gender_child2 gender_child3
1         30 1998-11-26 2000-01-29         NA             1             2            NA
2         27 1996-06-22         NA         NA             2            NA            NA
3         26 2002-07-11 2004-04-05 2007-09-02             2             2             1
4         32 2004-10-10 2009-08-27 2012-07-21             1             1             1
5         29 2000-12-05 2005-02-28         NA             2             1            NA")

family
```
And want to reshape the data.table to be like this:
```{r}
#     family_id age_mother  child        dob gender
#         <int>      <int> <char>     <char> <char>
#  1:         1         30 child1 1998-11-26      1
#  2:         1         30 child2 2000-01-29      2
#  3:         1         30 child3       <NA>   <NA>
#  4:         2         27 child1 1996-06-22      2
#  5:         2         27 child2       <NA>   <NA>
#  6:         2         27 child3       <NA>   <NA>
#  7:         3         26 child1 2002-07-11      2
#  8:         3         26 child2 2004-04-05      2
#  9:         3         26 child3 2007-09-02      1
# 10:         4         32 child1 2004-10-10      1
# 11:         4         32 child2 2009-08-27      1
# 12:         4         32 child3 2012-07-21      1
# 13:         5         29 child1 2000-12-05      2
# 14:         5         29 child2 2005-02-28      1
# 15:         5         29 child3       <NA>   <NA>
```

The `data.table::dcast` and `tidyr::pivot_longer` could transfer it in one step, however, not so easy to understand. Here we'll do it step by step to see what actually happens in this transfer.

```{r,warning=FALSE}
family %>% 
  longer_dt(1:2) %>% 
  separate_dt("name",into = c("class","child")) %>% 
  wider_dt(-"class|value",
           name = "class",
           value = "value")
```

In such a process, we could find that we actually get a longer table, then separate it, and wider it later. *tidyfst* is not going to support the complicated transfer in one step, because it might be easier to implement, but much harder to understand 3 procedures in 1 step. If you still prefer that way, use `data.table::dcast` and `tidyr::pivot_longer` instead.




