---
title: "Example 2: Join tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 2: Join tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
  This post has referred to a vignette from `dplyr`, you can find it in <https://dplyr.tidyverse.org/articles/two-table.html>. We'll try to display how to join data tables in this vignette. First, load the packages we need and get some data.
```{r setup}
library(tidyfst)
library(nycflights13)

flights2 <- flights %>% 
  select_dt(year,month,day, hour, origin, dest, tailnum, carrier)

```
  Do a left join with a simple:
```{r}
flights2 %>% 
  left_join_dt(airlines)
```
  
## Controlling how the tables are matched
  Join works the same as `dplyr`:
```{r}
flights2 %>% left_join_dt(weather)
flights2 %>% left_join_dt(planes, by = "tailnum")
flights2 %>% left_join_dt(airports, c("dest" = "faa"))
flights2 %>% left_join_dt(airports, c("origin" = "faa"))
```

## Types of join

```{r}
df1 <- data.table(x = c(1, 2), y = 2:1)
df2 <- data.table(x = c(1, 3), a = 10, b = "a")

df1 %>% inner_join_dt(df2) 
df1 %>% left_join_dt(df2)
df1 %>% right_join_dt(df2)
df1 %>% full_join_dt(df2)

```
  If all you have is a data.frame or tibble, you have no need to change the format. Feed the data directly:
```{r}
df1 <- data.frame(x = c(1, 1, 2), y = 1:3)
df2 <- data.frame(x = c(1, 1, 2), z = c("a", "b", "a"))

df1 %>% left_join_dt(df2)
```
  The "_dt" suffix should remind you that this is backed up by `data.table` and will always return a data.table in the end.

## Filtering joins
  Filtering joins have also been supported in `tidyfst`.
```{r}
flights %>% 
  anti_join_dt(planes, by = "tailnum") %>% 
  count_dt(tailnum, sort = TRUE)
```
  Other examples (`semi_join_dt()` and `anti_join_dt()` never duplicate; they only ever remove observations.):
```{r}
df1 <- data.frame(x = c(1, 1, 3, 4), y = 1:4)
df2 <- data.frame(x = c(1, 1, 2), z = c("a", "b", "a"))

# Four rows to start with:
df1 %>% nrow()

# And we get four rows after the join
df1 %>% inner_join_dt(df2, by = "x") %>% nrow()

# But only two rows actually match
df1 %>% semi_join_dt(df2, by = "x") %>% nrow()
```
  
## Set operations
  For set operations, wrap `data.table`'s function directly, but the functions will automatically turn any data.frame into data.table. Examples are listed as below:
  
```{r}
x = iris[c(2,3,3,4),]
x2 = iris[2:4,]
y = iris[c(3:5),]

intersect_dt(x, y)            # intersect
intersect_dt(x, y, all=TRUE)  # intersect all
setdiff_dt(x, y)              # except
setdiff_dt(x, y, all=TRUE)    # except all
union_dt(x, y)                # union
union_dt(x, y, all=TRUE)      # union all
setequal_dt(x, x2, all=FALSE) # setequal
setequal_dt(x, x2)     
```
  
For more details, just find the help from `data.table` using `?setops`.

