---
title: "Example 6: Dt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example6_dt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
  For absolute physical speed, use *data.table* directly. While the learning curve might be longer, the improvement of computation performance pays off if you are dealing with large datasets frequently. There are several ways to cut into *data.table* syntax to gain higher performance in *tidyfst*. A convenient way is to use the `DT[I,J,BY]` syntax after the pipe(`%>%`).
```{r setup}
library(tidyfst)
iris %>% 
  as_dt()%>%  #coerce a data.frame to data.table
  .[,.SD[1],by = Species]

```
  This syntax is not so consistent with the tidy syntax, therefore `in_dt` is also designed for the short cut to *data.table* method, which could be used as:
```{r}
iris %>% 
  in_dt(,.SD[1],by = Species)
```
  
  `in_dt` follows the basic principals of *tidyfst*, which include: (1) Never use in place replacement. Therefore, the in place functions like `:=` will still return the results. (2) Always recieves a data frame (data.frame/tibble/data.table) and returns a data.table. This means you don't have to write `as.data.table` or `as_dt` all the time as long as you are working on data frames in R.
 
  
