---
title: "Example 3: Reshape"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example3_reshape}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
  When I design `longer_dt` and `wider_dt`, I could find the `pivot_longer` and `pivot_wider` in `tidyr` and `melt` and `dcast` in `data.table`. Still, designing this API is not easy, as my goal is to let users use it with least pain. Here we would try to reproduce the results in the vignette of `tidyr`(<https://tidyr.tidyverse.org/articles/pivot.html>). First load the packages:
```{r setup}
library(tidyfst)
library(tidyr)
```

## Longer
  First inspect the data:
```{r}
relig_income
```
  In `tidyr`, to get the longer format you need:
```{r,eval=FALSE}
relig_income %>% 
  pivot_longer(-religion, names_to = "income", values_to = "count")
```
  In `tidyfst`, we have:
```{r,warning=FALSE}
relig_income %>% 
  longer_dt("religion",gathered_name = "income",gathered_value = "count")
```
  Another example from `tidyr`:
```{r,eval=FALSE}
billboard

# tidyr way:
 billboard %>%
   pivot_longer(
     cols = starts_with("wk"),
     names_to = "week",
     values_to = "rank",
     values_drop_na = TRUE
   )

# tidyfst way:
billboard %>% 
  longer_dt("wk",negate = TRUE,  
            gathered_name = "week",
            gathered_value = "rank",
            na.rm = TRUE
            )
# regex should select group_to_keep, negate could reverse that
```
A warning would could come out because the merging column has different data types and do the coercion automatically.
    
## Wider

```{r}
## data
fish_encounters

## tidyr way:
fish_encounters %>% 
  pivot_wider(names_from = station, values_from = seen)

## tidyfst way:
fish_encounters %>% 
  wider_dt(name_to_spread = "station",value_to_spread = "seen")
```
    
  If you want to fill with 0s, use:
```{r}
fish_encounters %>% 
  wider_dt(name_to_spread = "station",value_to_spread = "seen",fill = 0)
```
  Note that the parameter of `name_to_spread` and `value_to_spread` should always be provided and should be explicit called (with the parameter names attached).



