---
title: "Eample 1: Basic usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example1_intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette has referred to `dplyr`'s vignette in <https://dplyr.tidyverse.org/articles/dplyr.html>. We'll try to reproduce all the results. First load the needed packages.

```{r}
library(tidyfst)
library(nycflights13)

data.table(flights)
```

## Filter rows with `filter_dt()`
```{r}
filter_dt(flights, month == 1, day == 1)
```

## Arrange rows with `arrange_dt()`
```{r}
arrange_dt(flights, year, month, day)
```

  Use `-` (minus symbol) to order a column in descending order:
```{r}
arrange_dt(flights, -arr_delay)
```

## Select columns with `select_dt()`

```{r}
select_dt(flights, year, month, day)
```

  `select_dt(flights, year:day)` and `select_dt(flights, -(year:day))` are not supported. But I have added a feature to help select with regular expression, which means you can:
```{r}
select_dt(flights, "^dep")
```
  The rename process is almost the same as that in `dplyr`:
```{r}
select_dt(flights, tail_num = tailnum)
rename_dt(flights, tail_num = tailnum)
```
  
## Add new columns with `mutate_dt()`
```{r}
mutate_dt(flights,
  gain = arr_delay - dep_delay,
  speed = distance / air_time * 60
)
```
  
  However, if you just create the column, please split them. The following codes would not work:
```{r,eval=FALSE}
mutate_dt(flights,
  gain = arr_delay - dep_delay,
  gain_per_hour = gain / (air_time / 60)
)
```
  Instead, use:
```{r}
mutate_dt(flights,gain = arr_delay - dep_delay) %>%
  mutate_dt(gain_per_hour = gain / (air_time / 60))
```
  
  If you only want to keep the new variables, use `transmute_dt()`:
```{r}
transmute_dt(flights,
  gain = arr_delay - dep_delay
)
```
  
## Summarise values with `summarise_dt()`
```{r}
summarise_dt(flights,
  delay = mean(dep_delay, na.rm = TRUE)
)
```

## Randomly sample rows with `sample_n_dt()` and `sample_frac_dt()`
```{r}
sample_n_dt(flights, 10)
sample_frac_dt(flights, 0.01)
```

## Grouped operations
  For the below `dplyr` codes:
```{r,eval=FALSE}
by_tailnum <- group_by(flights, tailnum)
delay <- summarise(by_tailnum,
  count = n(),
  dist = mean(distance, na.rm = TRUE),
  delay = mean(arr_delay, na.rm = TRUE))
delay <- filter(delay, count > 20, dist < 2000)
```
  We could get it via:
```{r}
flights %>% 
  summarise_dt( count = .N,
  dist = mean(distance, na.rm = TRUE),
  delay = mean(arr_delay, na.rm = TRUE),by = tailnum)
```
  `summarise_dt` (or `summarize_dt`) has a parameter "by", you can specify the group.
  We could find the number of planes and the number of flights that go to each possible destination:
```{r}
# the dplyr syntax:
# destinations <- group_by(flights, dest)
# summarise(destinations,
#   planes = n_distinct(tailnum),
#   flights = n()
# )

summarise_dt(flights,planes = uniqueN(tailnum),flights = .N,by = dest) %>% 
  arrange_dt(dest)

```
  If you need to group by many variables, use:
```{r}
# the dplyr syntax:
# daily <- group_by(flights, year, month, day)
# (per_day   <- summarise(daily, flights = n()))

flights %>% 
  summarise_dt(by = .(year,month,day),flights = .N)

# (per_month <- summarise(per_day, flights = sum(flights)))
flights %>% 
  summarise_dt(by = .(year,month,day),flights = .N) %>% 
  summarise_dt(by = .(year,month),flights = sum(flights))

# (per_year  <- summarise(per_month, flights = sum(flights)))
flights %>% 
  summarise_dt(by = .(year,month,day),flights = .N) %>% 
  summarise_dt(by = .(year,month),flights = sum(flights)) %>% 
  summarise_dt(by = .(year),flights = sum(flights))
```
  


