---
title: "Eample 5: Tibble"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example5_tibble}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
  `tibble` is the name of a package, as well as a enhanced data.frame structure, which provides user lots of information automatically. I admire this design, and used to think of adding `as_tibble` to every end of function. This is, however, inefficient for data manipulation, which violates the principal of "fast". But thanks to Houyun, he taught me a way to adjust the `print.data.frame` function, which I finally use for `pring.data.table` for an alternative way of printing data.table. Now let's try it with the function `show_tibble`:
```{r setup}
library(tidyfst)

iris %>% count_dt(Species) -> a
class(a)

# turn on
show_tibble(TRUE)
a
class(a)

# shut off
show_tibble(FALSE)
a
class(a)
```

  We can find that the structure of variable `a` never changes, but the printing form of it could be altered to tibble or back to data.table using `show_tibble` function. You can choose the one you preferred.
  
  
